Interactive Checks for Coordination Avoidance

### 2019-08-20 doc/vldb_slides/sections/bank_example.tex
Progress towards slide revisions.

\newcommand{\bankdistance}{4}
\tikzstyle{bank}=[
  draw,
  minimum width=1.5cm,
  minimum height=1cm,
  font=\huge
]
\tikzstyle{op}=[
  -latex,
  ultra thick
]

\newcommand{\drawbanks}[6]{{
  \newcommand{\banka}{#1}
  \newcommand{\bankb}{#2}
  \newcommand{\bankaa}{#3}
  \newcommand{\bankbb}{#4}
  \newcommand{\bankaaa}{#5}
  \newcommand{\bankbbb}{#6}

  \node[bank, label=above:{Replica $A$}] (a1) at (0,              0) {\banka};
  \node[bank                           ] (a2) at (0,             -3) {\bankaa};
  \node[bank                           ] (a3) at (0,             -6) {\bankaaa};
  \node[bank, label=above:{Replica $B$}] (b1) at (\bankdistance,  0) {\bankb};
  \node[bank                           ] (b2) at (\bankdistance, -3) {\bankbb};
  \node[bank                           ] (b3) at (\bankdistance, -6) {\bankbbb};

  \draw[op] (a1) -- (a2);
  \draw[op] (a2) -- (a3);
  \draw[op] (b1) -- (b2);
  \draw[op] (b2) -- (b3);
}}

\newcommand{\zigzag}[2]{{
  \newcommand{\banka}{#1}
  \newcommand{\bankb}{#2}

  \draw[red, op] ($(\banka) + (0, -0.7)$) -- ($(\bankb) + (0, -1.0)$);
  \draw[red, op] ($(\bankb) + (0, -1.0)$) -- ($(\banka) + (0, -1.3)$);
  \draw[red, op] ($(\banka) + (0, -1.3)$) -- ($(\bankb) + (0, -1.6)$);
  \draw[red, op] ($(\bankb) + (0, -1.6)$) -- ($(\banka) + (0, -1.9)$);
}}

@@

  \note{%
    We'll begin with an example.

    Let's say we want to replicate the database shown here. This database is
    the world's simplest database. It contains only a single number, which
    represents my bank account balance. We decide to replicate this database
    across two machines, replica A and replica B.

    Users can issue transactions to the database, and every transacation is
    either a deposit or a withdrawal. And, I have one critical invariant that I
    want to maintain over my database. And that invariant is that my bank
    account balance should never be negative. My balance must be non-negative
    at all times.
  }
\end{frame}

\begin{frame}{Coordinating}
  \begin{center}
    \begin{tikzpicture}
      \drawbanks{\$50}{\$50}
                {\$60}{\$60}
                {\$80}{\$80}

      \draw[op] ($(a1) + (-2, 0)$) -- (a1) node[midway, above] {$+\$10$};
      \draw[op] ($(b1) + (+2, 0)$) -- (b1) node[midway, above] {$+\$20$};
      \draw[op] (a2) -- ($(a2) + (-2, 0)$) node[midway, above] {OK};
      \draw[op] (b3) -- ($(b3) + (+2, 0)$) node[midway, above] {OK};

      \zigzag{a1}{b1}
      \zigzag{a2}{b2}
    \end{tikzpicture}
  \end{center}

  \note{%
    One way to replicate our database is by using coordination to implement
    strong consistency. For example, imagine a user issues a deposit of 10
    dollars to replica A, and simultaneously, a user issues a deposit of 20
    dollars to replica B. The two replicas can coordinate, for example by
    running Paxos, and agree on a single total order in which to execute the
    two transactions. Here, the replicas agree to deposit 10 dollars and then
    deposit 20 dollars. Both transactions succeed.
  }
\end{frame}

\begin{frame}{Coordinating}
  \begin{center}
    \begin{tikzpicture}
      \drawbanks{\$50}{\$50}
                {\$20}{\$20}
                {\$20}{\$20}

      \draw[op] ($(a1) + (-2, 0)$) -- (a1) node[midway, above] {$-\$30$};
      \draw[op] ($(b1) + (+2, 0)$) -- (b1) node[midway, above] {$-\$40$};
      \draw[op] (a2) -- ($(a2) + (-2, 0)$) node[midway, above] {OK};
      \draw[op] (b3) -- ($(b3) + (+2, 0)$) node[midway, above] {NO};

      \zigzag{a1}{b1}
      \zigzag{a2}{b2}
    \end{tikzpicture}
  \end{center}

  \note{%
    Now, assume two clients issue concurrent withdrawal transactions of 30
    dollars and 40 dollars. Again, the two replicas coordinate and agree an a
    total order in which to execute transactions.

    They first withdraw 30 dollars, bringing my bank account balance to 20
    dollars. They then try to withdraw of 40 dollars but abort the transaction
    because I only have 40 dollars left. And while the transaction failed,
    you'll note that we never violated our application invariant.
  }
\end{frame}

\begin{frame}{Avoiding Coordination}
  \begin{center}
    \begin{tikzpicture}
      \drawbanks{\$50}{\$50}
                {\$60}{\$70}
                {\$80}{\$80}

      \draw[op] ($(a1) + (-2, 0)$) -- (a1) node[midway, above] {$+\$10$};
      \draw[op] ($(b1) + (+2, 0)$) -- (b1) node[midway, above] {$+\$20$};
      \draw[op] (a2) -- ($(a2) + (-2, 0)$) node[midway, above] {OK};
      \draw[op] (b2) -- ($(b2) + (+2, 0)$) node[midway, above] {OK};

      \draw[dashed, op] (a2) -- (b3) node[near start, sloped, above] {$+\$10$};
      \draw[dashed, op] (b2) -- (a3) node[near start, sloped, above] {$+\$20$};
    \end{tikzpicture}
  \end{center}

  \note{%
    Alternatively, instead of implementing our bank with strong consistency, we
    could implement it with weak consistency and avoid coordination altogether.

    Now, when two clients issue concurrent deposit transactions, the two
    replicas execute the transactions immediately without consulting one
    another. And every once in a while, the two replicas communicate and fill
    each other in on the transactions they missed. Notice that even though we
    didn't perform any coordination, the two replicas never violate my critical
    invariant of a non-negative balance.
  }
\end{frame}

\begin{frame}{Avoiding Coordination}
  \begin{center}
    \begin{tikzpicture}
      \drawbanks{\$50}{\$50}
                {\$20}{\$10}
                {-\$20}{-\$20}

      \draw[op] ($(a1) + (-2, 0)$) -- (a1) node[midway, above] {$-\$30$};
      \draw[op] ($(b1) + (+2, 0)$) -- (b1) node[midway, above] {$-\$40$};
      \draw[op] (a2) -- ($(a2) + (-2, 0)$) node[midway, above] {OK};
      \draw[op] (b2) -- ($(b2) + (+2, 0)$) node[midway, above] {OK};

      \draw[dashed, op] (a2) -- (b3) node[near start, sloped, above] {$-\$30$};
      \draw[dashed, op] (b2) -- (a3) node[near start, sloped, above] {$-\$40$};
    \end{tikzpicture}
  \end{center}

  \note{%
    However, let's look what happens when two clients issue concurrent
    withdrawals. Both withdrawals are executed locally and both succeed.
    However, when the two replicas exchange transactions with one another, they
    both enter an invariant violating state of negative twenty dollars!
  }
\end{frame}

\begin{frame}
  \begin{center}
    \Huge
    Deposits don't require coordination to maintain invariants, but withdrawals
    do.
  \end{center}

  \note{%
    We can learn a lot from these examples.

    First, if we implement our bank with strong consistency, we never violate
    our invariant.

    Second, we can implement deposit transactions without any coordination and
    we're still guaranteed to never violate the invariant.

    Third, withdrawal transactions, unlike deposit transactions, do require
    coordination to avoid violating the invariant.

    At a very high level, invariant confluence tries to capture exactly which
    transactions require coordination to maintain invariants and which do not.
    In this example, deposit transactions are invariant confluent and thus do
    not require coordination while withdrawal transactions are not invariant
    confluent and thus do require coordination.
  }
\end{frame}

>>>>

\newcommand{\bankdistance}{4}
\tikzstyle{bank}=[draw, minimum width=1.5cm, minimum height=1cm, font=\huge]
\tikzstyle{op}=[-latex, ultra thick]

\newcommand{\zigzag}[2]{{
  \newcommand{\banka}{#1}
  \newcommand{\bankb}{#2}
  \draw[gray, op] ($(\banka) + (0, -0.7)$) -- ($(\bankb) + (0, -1.0)$);
  \draw[gray, op] ($(\bankb) + (0, -1.0)$) -- ($(\banka) + (0, -1.3)$);
  \draw[gray, op] ($(\banka) + (0, -1.3)$) -- ($(\bankb) + (0, -1.6)$);
  \draw[gray, op] ($(\bankb) + (0, -1.6)$) -- ($(\banka) + (0, -1.9)$);
}}

@@

  \note{%
    We'll begin with an example. \\[12pt]

    Let's say we have an object that we want to replicate. The object can be
    something complex like a database or a key-value store, but to keep things
    simple, let's say we want to replicate a single integer that represents my
    bank account balance. We decide to replicate the integer across two
    machines, replica A and replica B. \\[12pt]

    Users can issue transactions to our replicate object, and every
    transacation is either a deposit or a withdrawal. And, I have one critical
    invariant that I want to maintain at all times. And that invariant is that
    my bank account balance should never be negative. \\[12pt]
  }
\end{frame}

\begin{frame}{Coordinating}
  \begin{center}
    \begin{tikzpicture}
      \node[bank, label=above:{Replica $A$}] (a1) at (0, 0) {\$50};
      \node[bank, label=above:{Replica $B$}] (b1) at (\bankdistance,  0) {\$50};
      \draw[op] ($(a1) + (-2, 0)$) -- (a1) node[midway, above] {$+\$10$};
      \draw[op] ($(b1) + (+2, 0)$) -- (b1) node[midway, above] {$+\$20$};
      \pause

      \zigzag{a1}{b1}
      \node[bank] (a2) at (0, -3) {\$60};
      \node[bank] (b2) at (\bankdistance, -3) {\$60};
      \draw[op] (a1) to node[left] {$+\$10$} (a2);
      \draw[op] (b1) to node[right] {$+\$10$} (b2);
      \pause

      \zigzag{a2}{b2}
      \node[bank] (a3) at (0, -6) {\$80};
      \node[bank] (b3) at (\bankdistance, -6) {\$80};
      \draw[op] (a2) to node[left] {$+\$20$} (a3);
      \draw[op] (b2) to node[right] {$+\$20$} (b3);
    \end{tikzpicture}
  \end{center}

  \note{%
    One way to replicate our object is by using coordination to implement
    strong consistency. For example, imagine a user issues a deposit of 10
    dollars to replica A, and simultaneously, another user issues a deposit of
    20 dollars to replica B. The two replicas can coordinate and agree on a
    single total order in which to execute the two transactions. Here, the
    replicas deposit 10 dollars and then deposit 20 dollars. Both transactions
    succeed and our invariant is maintained.
  }
\end{frame}

\begin{frame}{Coordinating}
  \begin{center}
    \begin{tikzpicture}
      \node[bank, label=above:{Replica $A$}] (a1) at (0, 0) {\$50};
      \node[bank, label=above:{Replica $B$}] (b1) at (\bankdistance,  0) {\$50};
      \draw[op] ($(a1) + (-2, 0)$) -- (a1) node[midway, above] {$-\$30$};
      \draw[op] ($(b1) + (+2, 0)$) -- (b1) node[midway, above] {$-\$40$};
      \pause

      \zigzag{a1}{b1}
      \node[bank] (a2) at (0, -3) {\$20};
      \node[bank] (b2) at (\bankdistance, -3) {\$20};
      \draw[op] (a1) to node[left] {$-\$30$} (a2);
      \draw[op] (b1) to node[right] {$-\$30$} (b2);
      \pause

      \zigzag{a2}{b2}
      \node[bank] (a3) at (0, -6) {\$20};
      \node[bank] (b3) at (\bankdistance, -6) {\$20};
      \draw[op] (a2) to node[left] {$-\$40$} (a3);
      \draw[op] (b2) to node[right] {$-\$40$} (b3);
    \end{tikzpicture}
  \end{center}

  \note{%
    Now, consider what happens when two clients issue concurrent withdrawal
    transactions of 30 dollars and 40 dollars. Again, the two replicas
    coordinate. They first withdraw 30 dollars, bringing my bank account
    balance to 20 dollars. They then try to withdraw of 40 dollars but abort
    the transaction because I only have 20 dollars left. Note that even though
    the withdrawal of 40 dollars failed, our invariant is always maintained.
  }
\end{frame}

\begin{frame}{Avoiding Coordination}
  \begin{center}
    \begin{tikzpicture}
      \node[bank, label=above:{Replica $A$}] (a1) at (0, 0) {\$50};
      \node[bank, label=above:{Replica $B$}] (b1) at (\bankdistance,  0) {\$50};
      \draw[op] ($(a1) + (-2, 0)$) -- (a1) node[midway, above] {$+\$10$};
      \draw[op] ($(b1) + (+2, 0)$) -- (b1) node[midway, above] {$+\$20$};
      \pause

      \node[bank] (a2) at (0, -3) {\$60};
      \node[bank] (b2) at (\bankdistance, -3) {\$70};
      \draw[op] (a1) to node[left] {$+\$10$} (a2);
      \draw[op] (b1) to node[right] {$+\$20$} (b2);
      \pause

      \node[bank] (a3) at (0, -6) {\$80};
      \node[bank] (b3) at (\bankdistance, -6) {\$80};
      \draw[op] (a2) to (a3);
      \draw[op] (b2) to (b3);
      \draw[dashed, op] (a2) -- (b3) node[near start, sloped, above] {$+\$10$};
      \draw[dashed, op] (b2) -- (a3) node[near start, sloped, above] {$+\$20$};
    \end{tikzpicture}
  \end{center}

  \note{%
    Alternatively, instead of implementing our bank with strong consistency, we
    can implement it with weak consistency and avoid coordination altogether.
    \\[12pt]

    Now, when two clients issue concurrent deposit transactions, the two
    replicas execute the transactions immediately without consulting one
    another. \\[12pt]

    And every once in a while, the two replicas communicate and fill each other
    in on the transactions they missed. Notice that even though we didn't
    perform any coordination, the two replicas never violate the invariant.
    \\[12pt]
  }
\end{frame}

\begin{frame}{Avoiding Coordination}
  \begin{center}
    \begin{tikzpicture}
      \node[bank, label=above:{Replica $A$}] (a1) at (0, 0) {\$50};
      \node[bank, label=above:{Replica $B$}] (b1) at (\bankdistance,  0) {\$50};
      \draw[op] ($(a1) + (-2, 0)$) -- (a1) node[midway, above] {$-\$30$};
      \draw[op] ($(b1) + (+2, 0)$) -- (b1) node[midway, above] {$-\$40$};
      \pause

      \node[bank] (a2) at (0, -3) {\$20};
      \node[bank] (b2) at (\bankdistance, -3) {\$10};
      \draw[op] (a1) to node[left] {$-\$30$} (a2);
      \draw[op] (b1) to node[right] {$-\$40$} (b2);
      \pause

      \node[bank] (a3) at (0, -6) {-\$20};
      \node[bank] (b3) at (\bankdistance, -6) {-\$20};
      \draw[op] (a2) to (a3);
      \draw[op] (b2) to (b3);
      \draw[dashed, op] (a2) -- (b3) node[near start, sloped, above] {$-\$30$};
      \draw[dashed, op] (b2) -- (a3) node[near start, sloped, above] {$-\$40$};
    \end{tikzpicture}
  \end{center}

  \note{%
    However, let's look at what happens when two clients issue concurrent
    withdrawals. Both withdrawals are executed locally and both succeed.
    However, when the two replicas exchange transactions with one another, they
    both enter an invariant violating state of negative twenty dollars!
  }
\end{frame}

\begin{frame}
  \Large

  \begin{center}
    \begin{enumerate}
      \item
        Strong consistency preserves the invariant.
      \pause \item
        Deposits don't require coordination.
      \pause \item
        Withdrawals do require coordination.
    \end{enumerate}
  \end{center}

  \note{%
    We can learn a lot from these examples. \\[12pt]

    First, if we implement our bank with strong consistency, we never violate
    our invariant. \\[12pt]

    Second, we can implement deposit transactions without any coordination and
    we're still guaranteed to never violate the invariant. \\[12pt]

    Third, withdrawal transactions, unlike deposit transactions, do require
    coordination to avoid violating the invariant. \\[12pt]

    At a very high level, invariant confluence tries to capture exactly which
    transactions require coordination to maintain invariants and which do not.
    In this example, deposit transactions are invariant confluent and thus do
    not require coordination, while withdrawal transactions not invariant
    confluent and thus do require coordination. \\[12pt]
  }
\end{frame}

### 2019-08-20 doc/vldb_slides/sections/four_wise_men.tex
Slimmed down talk.
Progress towards slide revisions.

\tikzstyle{pic}=[inner sep=0pt]
\tikzstyle{quote}=[fill=gray!30, inner sep=12pt]

\newcommand{\vogels}{
  \node[pic] (vogels) at (0, 0) {
    \includegraphics[
      width=0.22\textwidth,
      trim=0 75 0 25,
      clip
    ]{assets/vogels.jpg}
  };
}

\newcommand{\hoare}{
  \node[pic, right=0.2cm of vogels] (hoare) {
    \includegraphics[width=0.22\textwidth]{assets/hoare.jpg}
  };
}

\newcommand{\lamport}{
  \node[pic, right=0.2cm of hoare] (lamport) {
    \includegraphics[
      width=0.22\textwidth,
      trim=250 700 300 200,
      clip
    ]{assets/lamport.jpg}
  };
}

\newcommand{\brewer}{
  \node[pic, right=0.2cm of lamport] (brewer) {
    \includegraphics[
      width=0.22\textwidth,
      trim=100 300 100 100,
      clip
    ]{assets/brewer.jpg}
  };
}

\begin{frame}
  \note{%
    Hi everyone! Thanks so much for attending. I'm Michael, and today I'll be
    talking about ``Interactive Checks for Coordination Avoidance''. This is
    joint work with my advisor at UC Berkeley, Joe Hellerstein.

    Imagine you're an application developer and you're trying to decide what
    kind of consistency you'd like to use to replicate some of your data.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \node[quote, text width=0.8\textwidth, align=left] at (0, 3) {
        \Large\textit{You should use weak consistency. It's wicked fast!}
      };
      \path[fill=gray!30] (vogels.north) -- ++(0, 1) -- ++(0.5, 0) -- (vogels.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    Someone like Werner Vogels would probably tell you to use weak consistency.
    You can implement weak consistency super duper fast, and for many
    applications, weak consistency is good enough.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \hoare
      \node[quote, text width=0.8\textwidth, align=left] at (1, 3) {
        \Large\textit{What about the application invariants?!}
      };
      \path[fill=gray!30] (hoare.north) -- ++(0, 1.5) -- ++(1, 0) -- (hoare.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    But then Tony Hoare laments, ``no no no''. Hoare encourages us not to
    settle for weak consistency because a weakly consistent system is hard to
    reason about. It robs us of a fundamental way we use to reason about
    programs: application invariants. Even if every transaction in a weakly
    consistent system preserves application invariants, the whole system may
    not.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \hoare
      \lamport
      \node[quote, text width=0.8\textwidth, align=left] at (2, 3) {
        \Large\textit{Yeah, go with strong consistency! I have an algorithm for that :)}
      };
      \path[fill=gray!30] (lamport.north) -- ++(0, 1.5) -- ++(1, 0) -- (lamport.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    Leslie Lamport chimes in agreeingly. He encourages us to choose strong
    consistency. Strongly consistent systems are much easier to reason about.
    After all a strongly consistent system to end users looks exactly like a
    system that's not even distributed.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \hoare
      \lamport
      \brewer
      \node[quote, text width=0.8\textwidth, align=left] at (3, 3) {
        \Large\textit{Not so fast!}
      };
      \path[fill=gray!30] (brewer.north) -- ++(0, 1.5) -- ++(-1, 0) -- (brewer.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    ``Aha, not so fast'' counters Eric Brewer! Consistency and availability and
    partition tolerance, oh my. Surely, strong consistency comes at the price
    of availability and performance.

    So we're in a bit of a pickle. Weak consistency is too confusing and strong
    consistency is too heavy handed.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \node[pic] (bailis) at (0, 0) {\includegraphics[width=0.25\textwidth]{assets/bailis.jpg}};
      \node[quote, text width=0.8\textwidth, align=left] at (0, 3) {
        \Large\textit{?`Porque no los dos?}
      };
      \path[fill=gray!30] (bailis.north) -- ++(0, 1.5) -- ++(1, 0) -- (bailis.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    Thankfully, Peter Bailis comes to save the day and reminds us that perhaps
    we can have the best of both worlds. In 2014, Peter and some other smart
    folks including Alan defined invariant confluence as a condition under
    which we can replicate a piece of data without the need for strongly
    consistent coordination while maintaining the data's application
    invariants.

    Invariant confluence will be the topic of this talk. In particular, I'll
    define invariant confluence and then discuss how we can check invariant
    confluence automatically.
  }
\end{frame}

>>>>

\begin{frame}
  \tikzstyle{pic}=[inner sep=0pt]
  \tikzstyle{thought}=[cloud, draw, aspect=4]

  \begin{center}
    \begin{tikzpicture}[xscale=3, yscale=3]
      \node[pic] at (0, 0) {%
        \includegraphics[width=0.2\textwidth]{assets/person.jpg}
      };
      \pause

      \node[thought, aspect=1] at (-0.5, 0.5) {};
      \node[thought] at (-1, 1) {Weak Consistency?};
      \pause

      \node[thought, aspect=1] at (0.5, 0.5) {};
      \node[thought] at (1, 1) {Strong Consistency?};
      \pause

      \node[anchor=west] at (-1.75, 0.25) {%
        \textcolor{flatgreen}{\texttt{+}} Super fast
      };
      \node[anchor=west] at (-1.75, 0) {%
        \textcolor{flatred}{\texttt{-}} Hard to reason about
      };
      \pause

      \node[anchor=west] at (0.3, 0.25) {%
        \textcolor{flatgreen}{\texttt{+}} Easy to reason about
      };
      \node[anchor=west] at (0.3, 0) {%
        \textcolor{flatred}{\texttt{-}} Sacrifices availability
      };
    \end{tikzpicture}
  \end{center}

  \note{%
    Imagine you're an application developer and you have some data that you
    want to replicate. But, you're not sure whether to replicate the data with
    weak consistency or with strong consistency. \\[12pt]

    You can implement weak consistency super fast, but weakly consistent
    systems are hard to reason about. Strongly consistent systems are much
    easier to reason about, but they come at the cost of availability and
    performance. \\[12pt]
  }
\end{frame}

\begin{frame}
  \tikzstyle{pic}=[inner sep=0pt]
  \tikzstyle{quote}=[fill=gray!30, inner sep=12pt]

  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \node[pic] (bailis) at (0, 0) {%
        \includegraphics[width=0.25\textwidth]{assets/bailis.jpg}
      };
      \node[quote, text width=0.8\textwidth, align=left] at (0, 3) {
        \Large\textit{Why not have both?} Invariant confluence!
      };
      \path[fill=gray!30] (bailis.north) -- ++(0, 1.5) -- ++(1, 0) -- (bailis.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    Fortunately, Peter Bailis saves the day and reminds us of when in 2014, he
    and some other smart folks including Alan and my advisor Joe defined
    invariant confluence as a way to get the benefits of both weak and strong
    consistency.

    Invariant confluence will be the topic of this talk, so let's build some
    intuition on what invariant confluence is.
  }
\end{frame}
